---
title: "PamDx — Expression Analysis"
subtitle: "GTEx snRNA-seq + Cultured Fibroblast Heatmaps"
author: "Brouw"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = TRUE,
  warning = FALSE,
  fig.width  = 22,
  fig.height = 8
)
```

---

## Libraries

```{r libraries, message=FALSE}
library(Seurat)
library(BPCells)
library(dplyr)
library(tidyr)
library(magrittr)
library(stringr)
library(purrr)
library(tibble)
library(pheatmap)
library(grid)
library(data.table)
library(gridExtra)
```


## Configuration

All file paths are defined here. Edit this block only.

```{r config}
PATHS <- list(
  gtex_h5ad        = "C:/Users/brouw/OneDrive - Universiteit Utrecht/Master Bioinformatics/Minor internship/PamDx_TIM/Datasets/Expression_data/snRNA_seq/GTEx_8_tissues_snRNAseq_atlas_071421.public_obs.h5ad",
  gtex_parsed_dir  = "C:/Users/brouw/OneDrive - Universiteit Utrecht/Master Bioinformatics/Minor internship/PamDx_TIM/Datasets/Expression_data/snRNA_seq/GTEx_8_tissues_parsed",
  gtex_metadata    = "C:/Users/brouw/OneDrive - Universiteit Utrecht/Master Bioinformatics/Minor internship/PamDx_TIM/Datasets/Expression_data/snRNA_seq/GTEx_8_metadata.csv",
  gtex_genes       = "C:/Users/brouw/OneDrive - Universiteit Utrecht/Master Bioinformatics/Minor internship/PamDx_TIM/Datasets/Expression_data/snRNA_seq/GTEx_8_genes.csv",
  fibro_sample1    = "C:/Users/brouw/OneDrive - Universiteit Utrecht/Master Bioinformatics/Minor internship/PamDx_TIM/Datasets/Expression_data/cultured_fibroblasts/GSM8157769_HFF-50-CTR1.csv",
  fibro_sample2    = "C:/Users/brouw/OneDrive - Universiteit Utrecht/Master Bioinformatics/Minor internship/PamDx_TIM/Datasets/Expression_data/cultured_fibroblasts/GSM8157770_HFF-50-CTR2.csv",
  gtex_bulk_annot  =  "C:/Users/brouw/OneDrive - Universiteit Utrecht/Master Bioinformatics/Minor internship/PamDx_TIM/GTEx_Analysis_v11_Annotations_SampleAttributesDS.txt",
  gtex_bulk_counts = "C:/Users/brouw/Downloads/GTEx_Analysis_2026-05-19_v11_RNASeQCv2.4.3_gene_reads.gct/GTEx_Analysis_2025-08-22_v11_RNASeQCv2.4.3_gene_reads_non_lcm.gct",
  pamdx_lists_dir  = "C:/Users/brouw/OneDrive - Universiteit Utrecht/Master Bioinformatics/Minor internship/PamDx_TIM/Datasets/PamDX_lists",
  output_pdf_all   = "C:/Users/brouw/OneDrive - Universiteit Utrecht/Master Bioinformatics/Minor internship/PamDx_TIM/Datasets/PamDx_heatmaps.pdf",
  output_pdf_art   = "C:/Users/brouw/OneDrive - Universiteit Utrecht/Master Bioinformatics/Minor internship/PamDx_TIM/Datasets/PamDx_heatmaps_article1.pdf"
)
```

---

## Helper function — log(TP10K + 1) normalisation

All three expression datasets are normalised identically: counts are scaled to
10,000 per cell/sample (transcripts per 10k), then log~10~-transformed after
adding a pseudocount of 1.

```{r helper-norm}
log_tp10k_norm <- function(count_matrix) {
  lib_sizes <- colSums(count_matrix)
  tp10k     <- t(t(count_matrix) / lib_sizes) * 10000
  log10(tp10k + 1)
}
```

---

## 1. Load GTEx snRNA-seq dataset

```{r load-gtex-snrna}
GTEX_metadata <- read.csv(PATHS$gtex_metadata)
GTEX_genes    <- read.csv(PATHS$gtex_genes)

# Uncomment to write the BPCells directory the first time:
# parse.data <- open_matrix_anndata_hdf5(PATHS$gtex_h5ad)
# write_matrix_dir(mat = parse.data, dir = PATHS$gtex_parsed_dir)

GTEX_mat <- open_matrix_dir(dir = PATHS$gtex_parsed_dir)
```

---

## 2. Load cultured fibroblast dataset (GSE samples)

Two samples (HFF-50-CTR1 and HFF-50-CTR2) are joined on the gene-name column
and normalised to log(TP10K + 1).

```{r load-fibro-gse}
fibroblast_sample1 <- read.csv(PATHS$fibro_sample1)
fibroblast_sample2 <- read.csv(PATHS$fibro_sample2)

# Explicit join key to avoid silent row duplication
combined_fibroblast <- fibroblast_sample1 %>%
  left_join(fibroblast_sample2, by = "X", keep = FALSE)

rownames(combined_fibroblast) <- combined_fibroblast$X
combined_fibroblast %<>%
  dplyr::select(where(is.numeric))

# Normalise: log(TP10K + 1)
log10k_1 <- log_tp10k_norm(combined_fibroblast)
```

---

## 3. Build GTEx Seurat object and subset cardiac + skin fibroblasts

Cells are retained if they originate from heart tissue (all cell types) or
skin tissue (fibroblasts only). Skin fibroblasts are tagged with a `_Skin`
suffix to distinguish them from cardiac fibroblasts.

```{r build-gtex-seurat}
GTEX_object <- CreateSeuratObject(GTEX_mat, meta.data = GTEX_metadata)

cardiac_cells <- GTEX_object@meta.data[
  grepl("Heart", GTEX_object@meta.data$Tissue.Site.Detail, ignore.case = TRUE), ]

skin_fibro_cells <- GTEX_object@meta.data[
  grepl("skin", GTEX_object@meta.data$Tissue.Site.Detail, ignore.case = TRUE), ] %>%
  filter(Broad.cell.type == "Fibroblast")

subset_df <- rbind(cardiac_cells, skin_fibro_cells) %>%
  distinct(X)

GTEX_Seurat_subset <- subset(GTEX_object, cells = rownames(subset_df))
GTEX_Seurat_subset@assays$RNA@layers[["counts"]] <-
  as(GTEX_Seurat_subset@assays$RNA@layers[["counts"]], "dgCMatrix")

# Tag skin fibroblasts distinctly
GTEX_Seurat_subset@meta.data %<>%
  mutate(Broad.cell.type = if_else(
    str_starts(Tissue.Site.Detail, "Skin"),
    str_c(Broad.cell.type, "_Skin"),
    Broad.cell.type
  ))
```

---

## 4. Build cultured fibroblast Seurat object (GSE)

Data is already log(TP10K + 1) normalised — no further normalisation applied.

```{r build-fibro-seurat}
fibroblast_seurat <- CreateSeuratObject(log10k_1)
```

---

## 5. Load GTEx bulk RNA-seq cultured fibroblast samples

Samples annotated as `Cells - Cultured fibroblasts` in the GTEx v11 sample
attributes are extracted from the full gene read-count matrix and normalised
to log(TP10K + 1).

```{r load-gtex-bulk}
meta_bulk     <- fread(PATHS$gtex_bulk_annot)
fibro_samples <- meta_bulk[SMTSD == "Cells - Cultured fibroblasts", SAMPID]

GTEX_bulk <- fread(PATHS$gtex_bulk_counts, skip = 2)

fibro_samples_found   <- intersect(fibro_samples, colnames(GTEX_bulk))
fibro_samples_missing <- setdiff(fibro_samples,   colnames(GTEX_bulk))

message(sprintf(
  "GTEx bulk fibroblasts — found: %d  missing: %d",
  length(fibro_samples_found), length(fibro_samples_missing)
))

gene_names      <- GTEX_bulk$Description
GTEX_bulk_fibro <- as.matrix(GTEX_bulk[, fibro_samples_found, with = FALSE])
mode(GTEX_bulk_fibro) <- "numeric"
rownames(GTEX_bulk_fibro) <- gene_names

# Normalise: log(TP10K + 1)
GTEX_bulk_fibro_log <- log_tp10k_norm(GTEX_bulk_fibro)
rownames(GTEX_bulk_fibro_log) <- make.unique(rownames(GTEX_bulk_fibro_log))

fibroblasts_gtex <- CreateSeuratObject(GTEX_bulk_fibro_log)
```

---

## 6. Merge Seurat objects

The three datasets are merged into a single Seurat object. Missing
`Broad.cell.type` entries (from the cultured fibroblast datasets) are filled
with descriptive labels.

```{r merge-seurat}
fibro_gtex_seurat <- merge(GTEX_Seurat_subset, fibroblast_seurat, merge.data = TRUE)
fibro_gtex_seurat@meta.data %<>%
  mutate(Broad.cell.type = replace_na(Broad.cell.type, "Fibroblasts_cultured"))
fibro_gtex_seurat <- JoinLayers(fibro_gtex_seurat, layers = "counts")

fibro_gtex_seurat_fibroblast <- merge(fibro_gtex_seurat, fibroblasts_gtex, merge.data = TRUE)
fibro_gtex_seurat_fibroblast <- JoinLayers(fibro_gtex_seurat_fibroblast, layers = "counts")
fibro_gtex_seurat_fibroblast@meta.data %<>%
  mutate(Broad.cell.type = replace_na(Broad.cell.type, "Fibroblasts_cultured_GTEX"))
fibro_gtex_seurat_fibroblast <- JoinLayers(fibro_gtex_seurat_fibroblast)

analysis <- factor(fibro_gtex_seurat_fibroblast@meta.data[["Broad.cell.type"]])
message("Cell type levels: ", paste(levels(analysis), collapse = ", "))
message("NA cell types: ",    sum(is.na(analysis)))
```

---

## 7. Cell summary

```{r cell-summary}
cell_summary <- fibro_gtex_seurat@meta.data %>%
  filter(Broad.cell.type != "Fibroblasts_cultured") %>%
  group_by(Broad.cell.type) %>%
  summarise(
    n_cells         = n(),
    median_counts   = median(nCount_RNA,   na.rm = TRUE),
    median_features = median(nFeature_RNA, na.rm = TRUE)
  ) %>%
  arrange(desc(n_cells))

cell_summary
```

```{r overall-summary}
overall_summary <- cell_summary %>%
  summarise(
    n_celltypes    = n(),
    mean_n_cells   = mean(n_cells),
    sd_n_cells     = sd(n_cells),
    median_n_cells = median(n_cells),
    min_n_cells    = min(n_cells),
    max_n_cells    = max(n_cells)
  )

overall_summary
```

---

## 8. Load PamDx gene lists

```{r load-pamdx-lists}
pPs_all  <- read.csv(file.path(PATHS$pamdx_lists_dir, "pPs_all.csv"))
pPs_QCd  <- read.csv(file.path(PATHS$pamdx_lists_dir, "pPs_QCd.csv"))
STKs_all <- read.csv(file.path(PATHS$pamdx_lists_dir, "STKs_all.csv"))
STKs_QCd <- read.csv(file.path(PATHS$pamdx_lists_dir, "STKs_QCd.csv"))
```

```{r list-overview}
overview_lists <- list(
  pPs_all  = pPs_all,
  pPs_QCd  = pPs_QCd,
  STKs_all = STKs_all,
  STKs_QCd = STKs_QCd
)

imap_dfr(
  overview_lists,
  ~ tibble(
    list_name     = .y,
    n_rows        = nrow(.x),
    n_unique_rows = nrow(distinct(.x))
  )
)
```

---

## 9. Heatmap constants

Shared cell-type ordering, display labels, colour palette, and break points
used by every heatmap.

```{r heatmap-constants}
CELLTYPE_ORDER <- c(
  "Fibroblast-Skin",
  "Fibroblasts-cultured",
  "Fibroblasts-cultured-GTEX",
  "Fibroblast",
  "Myocyte (cardiac)",
  "Myocyte (cardiac, cytoplasmic)",
  "Adipocyte",
  "Endothelial cell (lymphatic)",
  "Endothelial cell (vascular)",
  "Immune (DC/macrophage)",
  "Immune (mast cell)",
  "Immune (NK cell)",
  "Immune (T cell)",
  "Pericyte/SMC",
  "Schwann cell"
)

ROW_LABELS <- c(
  "Fibroblast-Skin"                = "Fibroblast (skin)",
  "Fibroblasts-cultured"           = "Fibroblasts (cultured)",
  "Fibroblasts-cultured-GTEX"      = "Fibroblasts (cultured GTEX)",
  "Fibroblast"                     = "Heart - Fibroblast",
  "Myocyte (cardiac)"              = "Heart - Myocyte (cardiac)",
  "Myocyte (cardiac, cytoplasmic)" = "Heart - Myocyte (cardiac, cytoplasmic)",
  "Adipocyte"                      = "Heart - Adipocyte",
  "Endothelial cell (lymphatic)"   = "Heart - Endothelial cell (lymphatic)",
  "Endothelial cell (vascular)"    = "Heart - Endothelial cell (vascular)",
  "Immune (DC/macrophage)"         = "Heart - Immune (DC/macrophage)",
  "Immune (mast cell)"             = "Heart - Immune (mast cell)",
  "Immune (NK cell)"               = "Heart - Immune (NK cell)",
  "Immune (T cell)"                = "Heart - Immune (T cell)",
  "Pericyte/SMC"                   = "Heart - Pericyte/SMC",
  "Schwann cell"                   = "Heart - Schwann cell"
)

HEAT_COLORS <- colorRampPalette(c("black", "lightblue", "darkblue"))(100)

# Fine resolution near zero (sparse snRNA data), then linear to max_value
make_breaks <- function(max_value = 3) {
  c(
    seq(0,      0.003,     length.out = 50),
    seq(0.0031, max_value, length.out = 51)
  )
}
```

---

## 10. Core heatmap function

A single function replaces the two near-identical functions from the original
script. `order_by = "alphabetical"` or `"median"` controls gene ordering;
`split = TRUE` divides a wide heatmap into two stacked or side-by-side panels.

```{r heatmap-function}
make_pam_pheatmap <- function(
    seurat_obj,
    features,
    title,
    group_by     = "Broad.cell.type",
    assay_name   = "RNA",
    max_value    = 3,
    order_by     = c("alphabetical", "median"),
    decreasing   = TRUE,
    split        = FALSE,
    split_layout = c("side_by_side", "stacked"),
    cellwidth    = 10,
    cellheight   = 10,
    fontsize_row = 8,
    fontsize_col = 8,
    show_legend  = TRUE
) {
  order_by     <- match.arg(order_by)
  split_layout <- match.arg(split_layout)

  features <- unique(na.omit(features))
  features <- features[features %in% rownames(seurat_obj)]
  message(sprintf("[%s] Features found in object: %d", title, length(features)))


  avg_expr <- AverageExpression(
    seurat_obj,
    features = features,
    group.by = group_by,
    slot     = "counts"
  )

  mat <- as.matrix(avg_expr[[assay_name]])
  mat <- mat[features, , drop = FALSE]

  col_order <- CELLTYPE_ORDER[CELLTYPE_ORDER %in% colnames(mat)]
  mat       <- mat[, col_order, drop = FALSE]

  mat_t <- t(mat)
  storage.mode(mat_t) <- "numeric"
  mat_t[!is.finite(mat_t)] <- 0
  mat_t[mat_t < 0]         <- 0
  mat_t[mat_t > max_value] <- max_value

  if (order_by == "alphabetical") {
    mat_t <- mat_t[, sort(colnames(mat_t)), drop = FALSE]
  } else {
    med_score <- apply(mat_t, 2, median, na.rm = TRUE)
    mat_t     <- mat_t[, names(sort(med_score, decreasing = decreasing)), drop = FALSE]
  }

  breaks     <- make_breaks(max_value)
  labels_row <- ROW_LABELS[rownames(mat_t)]

  # ── Single panel ────────────────────────────────────────────────────────────
  if (!split) {
    p <- pheatmap::pheatmap(
      mat_t,
      color         = HEAT_COLORS,
      breaks        = breaks,
      cluster_rows  = FALSE,
      cluster_cols  = FALSE,
      show_rownames = TRUE,
      show_colnames = TRUE,
      labels_row    = labels_row,
      angle_col     = "90",
      cellwidth     = cellwidth,
      cellheight    = cellheight,
      fontsize_row  = fontsize_row,
      fontsize_col  = fontsize_col,
      fontsize      = 9,
      main          = title,
      border_color  = "grey90",
      legend        = show_legend,
      silent        = TRUE
    )
    grid::grid.draw(p$gtable)
    return(p)
  }

  # ── Split panel ─────────────────────────────────────────────────────────────
  mid     <- ceiling(ncol(mat_t) / 2)
  mat_t_1 <- mat_t[, 1:mid,                  drop = FALSE]
  mat_t_2 <- mat_t[, (mid + 1):ncol(mat_t),  drop = FALSE]

  make_panel <- function(m, suffix, legend) {
    pheatmap::pheatmap(
      m,
      color         = HEAT_COLORS,
      breaks        = breaks,
      cluster_rows  = FALSE,
      cluster_cols  = FALSE,
      show_rownames = TRUE,
      show_colnames = TRUE,
      labels_row    = ROW_LABELS[rownames(m)],
      angle_col     = "90",
      cellwidth     = cellwidth,
      cellheight    = cellheight,
      fontsize_row  = fontsize_row,
      fontsize_col  = fontsize_col,
      fontsize      = 9,
      main          = paste0(title, suffix),
      border_color  = "grey90",
      legend        = legend,
      silent        = TRUE
    )
  }

  p1 <- make_panel(mat_t_1, " (part 1)", show_legend)
  p2 <- make_panel(mat_t_2, " (part 2)", FALSE)

  ncols    <- if (split_layout == "side_by_side") 2L else 1L
  arranged <- gridExtra::arrangeGrob(p1$gtable, p2$gtable, ncol = ncols)
  grid::grid.draw(arranged)
  return(arranged)
}
```

---

## 11. Generate heatmaps

### 11a. Alphabetical gene ordering

```{r heatmap-pps-all-alpha, fig.cap="pPs-related genes, all tested — alphabetical"}
pPs_all_alphabetical <- make_pam_pheatmap(
  seurat_obj = fibro_gtex_seurat_fibroblast,
  features   = pPs_all$HGNC_genename,
  title      = "pPs-related genes, all tested — alphabetical",
  order_by   = "alphabetical"
)
```

```{r heatmap-pps-qcd-alpha, fig.cap="pPs-related genes, passed QC — alphabetical"}
pPs_QCd_alphabetical <- make_pam_pheatmap(
  seurat_obj = fibro_gtex_seurat_fibroblast,
  features   = pPs_QCd$HGNC.symbol,
  title      = "pPs-related genes, passed QC — alphabetical",
  order_by   = "alphabetical"
)
```

```{r heatmap-stks-all-alpha, fig.cap="inferred STKs (kinases), all tested — alphabetical"}
STKs_all_alphabetical <- make_pam_pheatmap(
  seurat_obj = fibro_gtex_seurat_fibroblast,
  features   = STKs_all$HGNC_genename,
  title      = "inferred STKs (kinases), all tested — alphabetical",
  order_by   = "alphabetical"
)
```

```{r heatmap-stks-qcd-alpha, fig.cap="inferred STKs (kinases), QCd — alphabetical"}
STKs_QCd_alphabetical <- make_pam_pheatmap(
  seurat_obj = fibro_gtex_seurat_fibroblast,
  features   = STKs_QCd$HGNC_genename,
  title      = "inferred STKs (kinases), QCd — alphabetical",
  order_by   = "alphabetical"
)
```

### 11b. Sorted by median expression

```{r heatmap-pps-all-median, fig.cap="pPs-related genes, all tested — sorted by median expression"}
pPs_all_median_sorted <- make_pam_pheatmap(
  seurat_obj = fibro_gtex_seurat_fibroblast,
  features   = pPs_all$HGNC_genename,
  title      = "pPs-related genes, all tested — sorted by median expression",
  order_by   = "median",
  decreasing = TRUE
)
```

```{r heatmap-pps-qcd-median, fig.cap="pPs-related genes, passed QC — sorted by median expression"}
pPs_QCd_median_sorted <- make_pam_pheatmap(
  seurat_obj = fibro_gtex_seurat_fibroblast,
  features   = pPs_QCd$HGNC.symbol,
  title      = "pPs-related genes, passed QC — sorted by median expression",
  order_by   = "median",
  decreasing = TRUE
)
```

```{r heatmap-stks-all-median, fig.cap="inferred STKs (kinases), all tested — sorted by median expression"}
STKs_all_median_sorted <- make_pam_pheatmap(
  seurat_obj = fibro_gtex_seurat_fibroblast,
  features   = STKs_all$HGNC_genename,
  title      = "inferred STKs (kinases), all tested — sorted by median expression",
  order_by   = "median",
  decreasing = TRUE
)
```

```{r heatmap-stks-qcd-median, fig.cap="inferred STKs (kinases), passed QC — sorted by median expression (split)"}
STKs_QCd_median_sorted <- make_pam_pheatmap(
  seurat_obj   = fibro_gtex_seurat_fibroblast,
  features     = STKs_QCd$HGNC_genename,
  title        = "inferred STKs (kinases), passed QC — sorted by median expression",
  order_by     = "median",
  decreasing   = TRUE,
  split        = TRUE,
  split_layout = "stacked"
)
```

---

## 12. Export PDFs

```{r export-pdf, results="hide"}
all_plots <- list(
  pPs_all_alphabetical,
  pPs_QCd_alphabetical,
  STKs_all_alphabetical,
  STKs_QCd_alphabetical,
  pPs_all_median_sorted,
  pPs_QCd_median_sorted,
  STKs_all_median_sorted,
  STKs_QCd_median_sorted
)

draw_plot <- function(p) {
  grid::grid.newpage()
  if ("gtable" %in% names(p)) grid::grid.draw(p$gtable) else grid::grid.draw(p)
}

# All heatmaps
pdf(PATHS$output_pdf_all, width = 22, height = 8)
invisible(lapply(all_plots, draw_plot))
dev.off()

# Article subset (QCd only)
# If this fails with 'unable to start device', run dev.off() until
# you see 'null device', then re-run this chunk.
grDevices::cairo_pdf(
  PATHS$output_pdf_art,
  width  = 22,
  height = 8,
  family = "Helvetica"
)
draw_plot(pPs_QCd_median_sorted)
draw_plot(STKs_QCd_median_sorted)
dev.off()

message("PDFs written to:\n  ", PATHS$output_pdf_all, "\n  ", PATHS$output_pdf_art)
```

---

## 13. Session info

```{r session-info}
sessionInfo()
```
